---
title: "Data 608 Assignment 3 Question One"
subtitle: "Interactive Visualizations in R"
author: "Stephen Haslett"
date: "9/29/2021"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rsconnect)
library(dplyr)
library(ggplot2)
library(shiny)
library(plotly)
```

### Assignment Instructions

I have provided you with data about mortality from all 50 states and the District of Columbia. Please access it at - https://github.com/charleyferrari/CUNY_DATA_608/tree/master/module3/data.

You are invited to gather more data from our provider, the CDC WONDER system, at https://wonder.cdc.gov. 

- This assignment must be done in R. It must be done using the ‘shiny’ package.
- It is recommended you use an R package that supports interactive graphing such as plotly, or vegalite, but this is not required.
- Your apps must be deployed, I won’t be accepting raw files. Luckily, you can pretty easily deploy apps with a free account at shinyapps.io.


```{r, eval=TRUE, message=FALSE, warning=FALSE}
# Import the mortality dataset into R. 
url <- "https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module3/data/cleaned-cdc-mortality-1999-2010-2.csv"
mortalityData <- read.csv(url, header = TRUE, stringsAsFactors = FALSE)
head(mortalityData)
```


### Question 1

As a researcher, you frequently compare mortality rates from particular causes across different States. You need a visualization that will let you see (for 2010 only) the crude mortality rate, across all States, from one cause (for example, Neoplasms, which are effectively cancers).

```{r question1, fig.height = 10, fig.width = 7, eval = TRUE, message = FALSE, warning = FALSE}
# Extract Neoplasm deaths for 2010 from the dataset.
mortalityData2010Neoplasms <- mortalityData %>%
  filter(., Year == '2010' & ICD.Chapter == 'Neoplasms') %>%
  arrange(desc(State), Crude.Rate)
```

Display a bar chart of 2010 crude mortality rates, across all States for neoplasms.
```{r question1BarChart, eval = TRUE, message = FALSE, warning = FALSE}
neoplasmCrudeMortalityChart <- mortalityData2010Neoplasms  %>%
  plot_ly(x = ~mortalityData2010Neoplasms$Crude.Rate,
          y = mortalityData2010Neoplasms$State,
          type = 'bar',
          orientation = 'h') %>%
  layout(title = list(text = '2010 Neoplasms Crude Mortality Rates By US State', font = list(size = 12), plot_bgcolor = '#e5ecf6'),
    xaxis = list(
      title = 'Crude Mortality Rate'
    ),
    yaxis = list(
      title = 'US State',
      categoryorder = 'array',
      categoryarray = rev(~State))
  )

subplot(neoplasmCrudeMortalityChart)   

```


Create a visualization that allows you to rank States by crude mortality for each cause of death.


```{r question1ShinyAppCode, eval = TRUE, message = FALSE, warning = FALSE}
# Define the Shiny UI.
ui <- fluidPage(
  headerPanel('US State Crude Mortality Rates by Cause of Death'),
  sidebarPanel(
    selectInput('Cause', 'Cause of Death', unique(mortalityData$ICD.Chapter),
                selected = 'Certain infectious and parasitic diseases',
                width = 'auto')
    
  ),
  mainPanel(
    htmlOutput(outputId = 'results'),
    plotOutput('CDRatePlot', height = 'auto')
  )
)

# Define the Shinny Server.
server <- shinyServer(function(input, output, session) {
  selectedData <- reactive({
    mortalityData  %>% filter(ICD.Chapter == input$Cause & Year == 2010)
  })
  
  output$selection <- renderText({
    paste('<i>Crude Mortality Rate For: </i>', input$Cause)
  })
  
  output$CDRatePlot <- renderPlot({
    ggplot(selectedData(), aes(x = reorder(State, -Crude.Rate), y = Crude.Rate)) +
      geom_col(fill = '#8B008B') +
      coord_flip() +
      geom_text(aes(label = Crude.Rate),
                size = 3,
                hjust = -0.2) +
      xlab('State') +
      ylab('Crude Mortality Rate') +
      theme(panel.background = element_blank())
  }, height = function() {
    session$clientData$output_CDRatePlot_width}
  )
})

# Render App One.
shinyApp(ui=ui, server=server)
```
