---
title: "Data608 Final Project"
author: "Stephen Haslett"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Overview

Your final project is to create a public visualization (static or interactive) using data relevant to a current policy, business, or justice issue. You may use any dataset you can find for this assignment, as long as it is either public or you have permission from the data’s owner/administrator to work with it and share it.
Recommended data sources are: governmental data, data provided by a non-profit/Nongovernmental organizations, and data available from large, semi-structured data sets (ie social networks, company financials, etc).

You must document each step of your data analysis process (excluding data acquisition) in code: this will include changing the format of the data and the creation of any images or interactive displays that are made.

You must also include a short (2-3 paragraph) write-up on the visualization. This write-up must include the following: the data source, what the parameters of the data set are (geography, timeframe, what the data points are, etc) what the data shows, and why it is important. Your proposal and your final visualization must be cleared by the instructor, and will be hosted on a public page (it will be available for the world to see).


## Assignment Objective


### Data Sources


#### Import the Data.

```{r dataImport, eval=TRUE, message=FALSE, warning=FALSE}

```


### Data Cleansing

```{r dataCleansing, eval=TRUE, message=FALSE, warning=FALSE}

```
