---
title: "Data608 Final Project"
author: "Stephen Haslett"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required libraries.
library(tidyverse)
library(caret)
library(kableExtra)
```

## Assignment Overview

Your final project is to create a public visualization (static or interactive) using data relevant to a current policy, business, or justice issue. You may use any dataset you can find for this assignment, as long as it is either public or you have permission from the data’s owner/administrator to work with it and share it.
Recommended data sources are: governmental data, data provided by a non-profit/Nongovernmental organizations, and data available from large, semi-structured data sets (ie social networks, company financials, etc).

You must document each step of your data analysis process (excluding data acquisition) in code: this will include changing the format of the data and the creation of any images or interactive displays that are made.

You must also include a short (2-3 paragraph) write-up on the visualization. This write-up must include the following: the data source, what the parameters of the data set are (geography, timeframe, what the data points are, etc) what the data shows, and why it is important. Your proposal and your final visualization must be cleared by the instructor, and will be hosted on a public page (it will be available for the world to see).


## Assignment Objective

For this assignment I would like to explore Worldwidew COVID rates. Although this is an exhaustively studied area these days, I feel that it fits the requirements of this assignment. This assignment aims to answer the following questions:



### Data Sources

To conduct this assignment, I will use the [Novel Corona Virus 2019](https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset) dataset that is available via **kaggle.com**. The dataset was taken from the [John Hopkins University COVID-19 Data Repository](https://github.com/CSSEGISandData/COVID-19).

The dataset consists of 6 different CSV files:

- **covid_19_data**  
    _The main dataset that will be used for this project. This file contains data on COVID cases_,  
    _deaths, and recoveries broken down by Country/Region, and Province/State were applicable._

- **time_series_covid_19_confirmed**  
    _Time series data on the number of confirmed cases worldwide._  
    _The file contains longitude and latitude data that can be used to plot cases on a map._  

- **time_series_covid_19_confirmed_US**  
    _Time series data on the number of confirmed cases in the US at the county level._  
    _This file will not be used in this project._

- **time_series_covid_19_deaths**  
    _Time series data on the number of COVID deaths worldwide._

- **time_series_covid_19_deaths_US**  
    _Time series data on the number of COVID deaths in the US at the county level._  
    _This file will not be used in this project._

- **time_series_covid_19_recovered**  
    _Time series data on the number of recovered COVID cases worlwide._

\ 

For the purposes of this assignment, I will be focusing on COVID at a worldwide level and will therefore ignore the US specific datasets. The datasets I will be using are as follows: **covid_19_data**, **time_series_covid_19_confirmed**, **time_series_covid_19_deaths**, and **time_series_covid_19_recovered**. 

\ 

### Import the Data

```{r dataImport, eval=TRUE, message=FALSE, warning=FALSE}
# Import the main dataset CSV file.
covid_data_main <- read.csv('./final_project_data/covid_19_data.csv')

# Import the confirmed cases time series CSV file.
covid_ts_confirmed_cases <- read.csv('./final_project_data/time_series_covid_19_confirmed.csv')

# Import the COVID deaths time series CSV file. 
covid_ts_deaths <- read.csv('./final_project_data/time_series_covid_19_deaths.csv')

# Import the COVID recoveries time series CSV file.
covid_ts_recovered <- read.csv('./final_project_data/time_series_covid_19_recovered.csv')
```

\ 

### Data Exploration and Cleansing

#### Main Dataset
```{r mainDataset, eval=TRUE, message=FALSE, warning=FALSE}
# Take a look at the structure of the main dataset.
head(covid_data_main) %>% kable() %>% kable_styling()
```


#### Confirmed Cases Time Series Dataset
```{r confirmedCases, eval=TRUE, message=FALSE, warning=FALSE}
```


#### COVID Deaths Time Series Dataset
```{r covidDeaths, eval=TRUE, message=FALSE, warning=FALSE}
```


#### COVID Recoveries Time Series Dataset
```{r covidRecoveries, eval=TRUE, message=FALSE, warning=FALSE}
```


